---
title: "Using Reticulate"
author: "Frida Gomam"
date: 2020-12-06
categories: ["R", "Python"]
tags: ["R", "Python", "reticulate"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

#Reticulate
Whenever you need R and Python to work together and share information, you can use reticulate. 

This is fairly easy!

When you insert a chunck (shortcut: Ctrl+Alt+I OR Command+Option+I), you will see `{r}`. That means that that chunk will be using R. You can change `{r}` to `{python}` to simply switch the chunk to a Python code chunk. 

First, you need to run `library(reticulate)` in the R code chunk to make R and Python communicate with each other!

```{r}
library(reticulate)
```

Then, you can run anything using R. For example, you can do this:

```{r}
#this is an R code chunk
fromR <- "this is from R"
```

You can have a Python code chunk to print `fromR` by putting `r.` in front of it. 
```{python}
#this is a Python code chunk
print(r.fromR)
```

You can do this vice versa!
```{python}
#this is a Python code chunk
fromPython = "this is from Python"
```

Instead of `r.`, use `py$` in an R chunk to bring something from the Python chunk. 
```{r}
#this is an R chunk
py$fromPython
```

Isn't this pretty cool?! You can do a lot more things using reticulate. I will post some more examples in the future!

